---
title: "R Notebook"
output: html_notebook
---
#Set seed
```{r}
set.seed(2025)
```

#Packages required for all models
```{r}
library("devtools") #For Bayesian
library("ggplot2") # For All
library("dplyr") #For All
library("MASS") # For Negative Binomial regression - Karlinsky and Kobak's
library("lmtest") #For 
library("mosaic")
library("car")
library("vroom")
library("tidyverse") #For Poisson regression
library("readr") 
library("data.table")
library("lubridate")
library("brms") #For Bayesian
library("cmdstanr") #For Bayesian
library("rstan") #For Bayesian
```

#Datasets used
```{r}
#Total Deaths Broken Down by Age and Gender
research1.df = read.csv("C:\\Users\\Alyssa Dollente\\Downloads\\DATA 501 L01\\Research Project\\Dataset - Provinces\\Deaths_1x1_BritishColumbia.csv")
research1.df
```


#(1) H2N2 "Asian flu" - Influenza 1 (1957-58)
```{r}
#For the first 3 models
influenza1.history = filter(research1.df, Year >= "1951" & Year <= "1956")
#influenza1.history

influenza1.timeline = filter(research1.df, Year >= "1957" & Year <= "1958")
#influenza1.timeline
```

#Traning and Validation data 
```{r}
# Split into training (1952-1956) and validation (1957-1958) sets
training_data <- influenza1.history
validation_data <- influenza1.timeline
```

#(1-c) Bayesian Regression Model
```{r}
#Bayesian Regression Model

# Fit Bayesian Poisson model with age group as a predictor
bayesian_model <- brm(
  as.integer(Total) ~ Year + Age, 
  family = poisson(),
  data = training_data,
  prior = c(prior(normal(0, 5), class = "b")),
  chains = 4, iter = 2000, warmup = 1000, cores = 4
)

# Predict expected deaths for validation set
predictions <- posterior_predict(bayesian_model, newdata = validation_data)
validation_data$expected_deaths <- apply(predictions, 2, mean)  # Mean of posterior samples

# Compute excess mortality for each age group
validation_data$excess_mortality <- validation_data$Total - validation_data$expected_deaths

# Summarize total excess mortality across all age groups per year
total_excess <- validation_data %>%
  group_by(Year) %>%
  summarize(excess_mortality = sum(excess_mortality), 
            observed_deaths = sum(Total), 
            expected_deaths = sum(expected_deaths), .groups = "drop")

#Adding the excess mortality values into the validation data
validation_data$excess_mortality_bayes_1957 <- validation_data$excess_mortality
excess_mortality_bayes_1957 <- validation_data$excess_mortality_bayes_1957

#Summary of excess deaths and the Bayesian Regression model
total_excess 
summary(bayesian_model)
```


```{r}
# Calculate quantiles for each year (across all draws in 'predictions')
lower_quantile <- apply(predictions, 2, function(x) quantile(x, probs = 0.025))  # Lower 2.5%
upper_quantile <- apply(predictions, 2, function(x) quantile(x, probs = 0.975))   # Upper 97.5%

# Map the quantiles to the 'total_excess' data based on 'year'
total_excess$lower <- lower_quantile[match(total_excess$Year, colnames(predictions))]
total_excess$upper <- upper_quantile[match(total_excess$Year, colnames(predictions))]

#Plot results
ggplot(total_excess, aes(x = Year)) +
  geom_line(aes(y = observed_deaths, color = "Observed Deaths"), linewidth = 1) +
  geom_line(aes(y = expected_deaths, color = "Expected Deaths"), linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = lower,
                  ymax = upper),
              fill = "blue", alpha = 0.2) +
  geom_col(aes(y = excess_mortality, fill = excess_mortality > 0), alpha = 0.5) +
  scale_color_manual(values = c("Observed Deaths" = "red", "Expected Deaths" = "blue")) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  scale_x_continuous(breaks = seq(min(total_excess$Year), max(total_excess$Year), by = 1)) +   
       labs(title = "Excess Mortality Estimation (Bayesian Poisson Model with Age Adjustment)",
       x = "Year", y = "Number of Deaths",
       color = "Legend", fill = "Excess Mortality") +
  theme_minimal()
```

#####Summary of excess mortality across all models
```{r}
# Summary of Excess Mortality across All Models

# Combine excess mortality results into a single data frame
excess_summary <- data.frame(
  Year = validation_data$Year,
  Bayesian = validation_data$excess_mortality_bayes_1957
)

# Summarize total excess mortality across years
excess_summary_total <- excess_summary %>%
  group_by(Year) %>%
  summarize(Bayesian = sum(Bayesian, na.rm = TRUE),
    .groups = "drop"
  )

print(excess_summary_total)

# Plot excess mortality comparisons
ggplot(excess_summary_total, aes(x = Year)) +
  geom_line(aes(y = Bayesian, color = "Bayesian"), linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Quasi-Poisson" = "blue", "Negative Binomial" = "green", "Bayesian" = "red")) +
  scale_x_continuous(breaks = seq(min(excess_summary_total$Year), max(excess_summary_total$Year), by = 1)) +
  labs(title = "Comparison of Excess Mortality Estimates Across Models",
       x = "Year", y = "Excess Mortality",
       color = "Model") +
  theme_minimal()
```

#(2) H3N2 "Hong Kong Flu" - Influenza 2 (1968-69)
```{r}
#For the first 3 models
influenza2.history = filter(research1.df, Year >= "1962" & Year <= "1967")
#influenza2.history

influenza2.timeline = filter(research1.df, Year >= "1968" & Year <= "1969")
influenza2.timeline
```

#Training and Validation data
```{r}
# Split into training (1962-1967) and validation (1968-1969) sets
training_data <- influenza2.history
validation_data <- influenza2.timeline
```

#(2-a) Quasi-Possion Regression Model
```{r}
#Quasi-Poisson regression model 

# Fit Poisson regression model
poisson_model <- glm(Total ~ Year + Age, family = quasipoisson, data = training_data)

# Predict expected deaths for 2020-2022
validation_data$expected_deaths <- predict(poisson_model, newdata = validation_data, type = "response")

# Compute excess mortality
validation_data$excess_mortality <- validation_data$Total - validation_data$expected_deaths

# Summarize total excess mortality across all age groups per year
total_excess <- validation_data %>%
  group_by(Year) %>%
  summarize(excess_mortality = sum(excess_mortality), 
            observed_deaths = sum(Total), 
            expected_deaths = sum(expected_deaths), .groups = "drop")

total_excess
```

```{r}
#Adding the excess mortality values into the validation data
validation_data$excess_mortality_qp_1968 <- validation_data$excess_mortality
excess_mortality_qp_1968 <- validation_data$excess_mortality_qp_1968

#Summary of excess deaths and the Quasi-Poisson model
total_excess
summary(poisson_model)
```

```{r}
# Plot results
ggplot(total_excess, aes(x = Year)) +
  geom_line(aes(y = observed_deaths, color = "Observed Deaths"), linewidth = 1) +
  geom_line(aes(y = expected_deaths, color = "Expected Deaths"), linetype = "dashed", linewidth = 1) +
  geom_col(aes(y = excess_mortality, fill = excess_mortality > 0), alpha = 0.5) +
  scale_color_manual(values = c("Observed Deaths" = "red", "Expected Deaths" = "blue")) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  scale_x_continuous(breaks = seq(min(total_excess$Year), max(total_excess$Year), by = 1)) + 
  labs(title = "Excess Mortality Estimation (Poisson Regression with Age Adjustment)",
       x = "Year", y = "Number of Deaths",
       color = "Legend", fill = "Excess Mortality") +
  theme_minimal()
```

#####Summary of excess mortality across all models
```{r}
# Summary of Excess Mortality across All Models

# Combine excess mortality results into a single data frame
excess_summary <- data.frame(
  Year = validation_data$Year,
  Quasi_Poisson = validation_data$excess_mortality_qp_1968
)

# Summarize total excess mortality across years
excess_summary_total <- excess_summary %>%
  group_by(Year) %>%
  summarize(
    Quasi_Poisson = sum(Quasi_Poisson, na.rm = TRUE),
    .groups = "drop"
  )

print(excess_summary_total)

# Plot excess mortality comparisons
ggplot(excess_summary_total, aes(x = Year)) +
  geom_line(aes(y = Quasi_Poisson, color = "Quasi-Poisson"), linewidth = 1) +
  scale_color_manual(values = c("Quasi-Poisson" = "blue", "Negative Binomial" = "green", "Bayesian" = "red")) +
  scale_x_continuous(breaks = seq(min(excess_summary_total$Year), max(excess_summary_total$Year), by = 1)) +
  labs(title = "Comparison of Excess Mortality Estimates Across Models",
       x = "Year", y = "Excess Mortality",
       color = "Model") +
  theme_minimal()
```

#(3) H1N1 "Swine Flu" - Influenza 2 (2009-10)
```{r}
#For the first 3 models
influenza3.history = filter(research1.df, Year >= "2003" & Year <= "2008")
#influenza3.history

influenza3.timeline = filter(research1.df, Year >= "2009" & Year <= "2010")
#influenza3.timeline
```

#Traning and Validation data
```{r}
# Split into training (2003-2008) and validation (2009-2010) sets
training_data <- influenza3.history
validation_data <- influenza3.timeline
```

#(3-c) Bayesian Regression Model
```{r}
#Bayesian Regression Model

# Fit Bayesian Poisson model with age group as a predictor
bayesian_model <- brm(
  as.integer(Total) ~ Year + Age, 
  family = poisson(),
  data = training_data,
  prior = c(prior(normal(0, 5), class = "b")),
  chains = 4, iter = 2000, warmup = 1000, cores = 4
)

# Predict expected deaths for validation set
predictions <- posterior_predict(bayesian_model, newdata = validation_data)
validation_data$expected_deaths <- apply(predictions, 2, mean)  # Mean of posterior samples

# Compute excess mortality for each age group
validation_data$excess_mortality <- validation_data$Total - validation_data$expected_deaths

# Summarize total excess mortality across all age groups per year
total_excess <- validation_data %>%
  group_by(Year) %>%
  summarize(excess_mortality = sum(excess_mortality), 
            observed_deaths = sum(Total), 
            expected_deaths = sum(expected_deaths), .groups = "drop")

# Calculate quantiles for each year (across all draws in 'predictions')
lower_quantile <- apply(predictions, 2, function(x) quantile(x, probs = 0.025))  # Lower 2.5%
upper_quantile <- apply(predictions, 2, function(x) quantile(x, probs = 0.975))   # Upper 97.5%

# Map the quantiles to the 'total_excess' data based on 'year'
total_excess$lower <- lower_quantile[match(total_excess$Year, colnames(predictions))]
total_excess$upper <- upper_quantile[match(total_excess$Year, colnames(predictions))]

#Adding the excess mortality values into the validation data
validation_data$excess_mortality_bayes_2009 <- validation_data$excess_mortality
excess_mortality_bayes_2009 <- validation_data$excess_mortality_bayes_2009

#Summary of excess deaths and the Bayesian Regression model
total_excess
summary(bayesian_model)
```

```{r}
#Plot results
ggplot(total_excess, aes(x = Year)) +
  geom_line(aes(y = observed_deaths, color = "Observed Deaths"), linewidth = 1) +
  geom_line(aes(y = expected_deaths, color = "Expected Deaths"), linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = lower,
                  ymax = upper),
              fill = "blue", alpha = 0.2) +
  geom_col(aes(y = excess_mortality, fill = excess_mortality > 0), alpha = 0.5) +
  scale_color_manual(values = c("Observed Deaths" = "red", "Expected Deaths" = "blue")) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  scale_x_continuous(breaks = seq(min(total_excess$Year), max(total_excess$Year), by = 1)) + 
  labs(title = "Excess Mortality Estimation (Bayesian Poisson Model with Age Adjustment)",
       x = "Year", y = "Number of Deaths",
       color = "Legend", fill = "Excess Mortality") +
  theme_minimal()
```

#####Summary of excess mortality across all models
```{r}
# Summary of Excess Mortality across All Models

# Combine excess mortality results into a single data frame
excess_summary <- data.frame(
  Year = validation_data$Year,
  Bayesian = validation_data$excess_mortality_bayes_2009
)

# Summarize total excess mortality across years
excess_summary_total <- excess_summary %>%
  group_by(Year) %>%
  summarize(
    Bayesian = sum(Bayesian, na.rm = TRUE),
    .groups = "drop"
  )

print(excess_summary_total)

# Plot excess mortality comparisons
ggplot(excess_summary_total, aes(x = Year)) +
  geom_line(aes(y = Bayesian, color = "Bayesian"), linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Quasi-Poisson" = "blue", "Negative Binomial" = "green", "Bayesian" = "red")) +
  scale_x_continuous(breaks = seq(min(excess_summary_total$Year), max(excess_summary_total$Year), by = 1)) +
  labs(title = "Comparison of Excess Mortality Estimates Across Models",
       x = "Year", y = "Excess Mortality",
       color = "Model") +
  theme_minimal()
```

#(4) COVID-19 (2020-22)
```{r}
#For the first 3 models
covid19.history = filter(research1.df, Year >= "2014" & Year <= "2019")
#covid19.history

covid19.timeline = filter(research1.df, Year >= "2020" & Year <= "2021") #change from the Canadian dataset since it only has mortality count up until 2021
#covid19.timeline
```

#Training and Validation data
```{r}
# Split into training (2014-2019) and validation (2020-2021) sets
training_data <- covid19.history
validation_data <- covid19.timeline
```

#(4-b) Karlinsky and Kobak's Regression model
```{r}
#Karlinsky and Kobak's Regression Model

# Ensure deaths are integers (needed for Poisson models)
training_data$Total <- as.integer(round(training_data$Total))
validation_data$Total <- as.integer(round(validation_data$Total))

# Fit Karlinsky & Kobak's model (Quasi-Poisson or Negative Binomial)
# If overdispersion is high, use Negative Binomial
model_nb <- glm.nb(Total ~ Year + Age, data = training_data)

# Predict expected deaths for validation set
validation_data$expected_deaths <- predict(model_nb, newdata = validation_data, type = "response")

# Compute excess mortality for each age group
validation_data$excess_mortality <- validation_data$Total - validation_data$expected_deaths

# Summarize total excess mortality across all age groups per year
total_excess <- validation_data %>%
  group_by(Year) %>%
  summarize(excess_mortality = sum(excess_mortality), 
            observed_deaths = sum(Total), 
            expected_deaths = sum(expected_deaths), .groups = "drop")

#Adding the excess mortality values into the validation data
validation_data$excess_mortality_nb_covid <- validation_data$excess_mortality
excess_mortality_nb_covid <- validation_data$excess_mortality_nb_covid


#Summary of excess deaths and the Karlinsky and Kobak's model
total_excess
summary(model_nb)
```

```{r}
# Plot results
ggplot(total_excess, aes(x = Year)) +
  geom_line(aes(y = observed_deaths, color = "Observed Deaths"), linewidth = 1) +
  geom_line(aes(y = expected_deaths, color = "Expected Deaths"), linetype = "dashed", linewidth = 1) +
  geom_col(aes(y = excess_mortality, fill = excess_mortality > 0), alpha = 0.5) +
  scale_color_manual(values = c("Observed Deaths" = "red", "Expected Deaths" = "blue")) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  scale_x_continuous(breaks = seq(min(total_excess$Year), max(total_excess$Year), by = 1)) + 
  labs(title = "COVID-19 Excess Mortality Estimation (2020-22)",
       color = "Legend", fill = "Excess Mortality") + theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal()
```

#####Summary of excess mortality across all models
```{r}
# Summary of Excess Mortality across All Models

# Combine excess mortality results into a single data frame
excess_summary <- data.frame(
  Year = validation_data$Year,
  Negative_Binomial = validation_data$excess_mortality_nb_covid
)

# Summarize total excess mortality across years
excess_summary_total <- excess_summary %>%
  group_by(Year) %>%
  summarize(
    Negative_Binomial = sum(Negative_Binomial, na.rm = TRUE),
    .groups = "drop"
  )

print(excess_summary_total)

# Plot excess mortality comparisons
ggplot(excess_summary_total, aes(x = Year)) +
  geom_line(aes(y = Negative_Binomial, color = "Negative Binomial"), linewidth = 1) +
  scale_color_manual(values = c("Quasi-Poisson" = "blue", "Negative Binomial" = "green", "Bayesian" = "red")) +
  scale_x_continuous(breaks = seq(min(excess_summary_total$Year), max(excess_summary_total$Year), by = 1)) +
  labs(title = "Comparison of Excess Mortality Estimates Across Models",
       x = "Year", y = "Excess Mortality",
       color = "Model") +
  theme_minimal()
```

#######Summary of excess mortality across all pandemics - all in one sheet
```{r}
#Aggregate all the pandemic names
pandemic_names <- c(
  rep("H2N2 Asian Flu", length(influenza1.timeline$Year)),
  rep("H3N2 Hong Kong Flu", length(influenza2.timeline$Year)),
  rep("H1N1 Swine Flu", length(influenza3.timeline$Year)),
  rep("COVID-19", length(covid19.timeline$Year))
)

#Combine all of the timeline into one list
pandemic_timeline <- c(influenza1.timeline$Year,influenza2.timeline$Year,influenza3.timeline$Year,covid19.timeline$Year)

#Combine all of the excess mortality into one list
excess_mortality_united <- c(excess_mortality_bayes_1957,excess_mortality_qp_1968,excess_mortality_bayes_2009,excess_mortality_nb_covid)

# Combine all excess mortality into one
excess_mortality_comparison <- data.frame(
  Year = pandemic_timeline,
  Age = validation_data$Age,
  Excess_Mortality = excess_mortality_united,
  Pandemic = pandemic_names 
)

# Print results
excess_mortality_comparison

# Save as a CSV file
write.csv(excess_mortality_comparison, "Pandemic_Excess_Mortality_BritishColumbia_i.csv", row.names = FALSE)
```

```{r}
library("openxlsx") #to create an excel file for excess mortality results
```


######Summary of excess mortality across all pandemics
```{r}
# Create a new workbook to store the results
wb <- createWorkbook()

# List of pandemics and the corresponding years and excess mortality variables
pandemic_data <- list(
  H2N2_Asian_Flu = list(years = c(1957, 1958), 
                        bayes = excess_mortality_bayes_1957, 
                        timeline = influenza1.timeline),
  
  H3N2_Hong_Kong_Flu = list(years = c(1968, 1969), 
                            qp = excess_mortality_qp_1968, 
                            timeline = influenza2.timeline),
  
  H1N1_Swine_Flu = list(years = c(2009, 2010), 
                        bayes = excess_mortality_bayes_2009, 
                        timeline = influenza3.timeline),
  
  COVID19 = list(years = c(2020, 2021), 
                 nb = excess_mortality_nb_covid, 
                 timeline = covid19.timeline)
)

pandemic_data
```


```{r}
# Loop through each pandemic
for (pandemic in names(pandemic_data)) {
  
  # Get the relevant excess mortality data and timeline for this pandemic
  pandemic_info <- pandemic_data[[pandemic]]
  years <- pandemic_info$years
  timeline <- pandemic_info$timeline
  
  # Determine the number of rows based on the length of the excess mortality data
  num_rows <- length(pandemic_info$qp)
  
  # Create an empty data frame to store results for this pandemic while attaching
  #original Year and Age from the validation dataset (timeline)
  pandemic_results <- data.frame(Year = timeline$Year, 
                                 Age = timeline$Age)
  
  # Add the excess mortality values for each model
  pandemic_results$Quasi_Poisson <- pandemic_info$qp
  pandemic_results$Negative_Binomial <- pandemic_info$nb
  pandemic_results$Bayesian <- pandemic_info$bayes
  
  # Create a simple sheet name (limit to the first 31 characters to avoid exceeding Excel's limit)
  sheet_name <- substr(pandemic, 1, 31)
  
  # Add this pandemic's data to a new sheet in the workbook
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, pandemic_results)
}

# Save the workbook as an Excel file
saveWorkbook(wb, "Pandemic_Excess_Mortality_BritishColumbia.xlsx", overwrite = TRUE)

cat("Excess mortality results saved as an Excel file with multiple sheets for each pandemic.\n")
```

